---
title: "Cold Tolerance Analysis"
author: "FiG-T"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: yes
    toc: yes
toc: TRUE
editor_options:
  markdown:
    wrap: 80
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries & setup

```{r libaries}
# select required libraries
library(readxl)
library(dplyr)
library(ggplot2)
```

```{r colour_scheme}

source("https://raw.githubusercontent.com/FiG-T/scripts/refs/heads/main/R/time_r/time_palettes_themes.R")
```

# Introduction

As part of the experimental evolution study looking at mitonuclear
incompatibilities (T.I.M.E) we assess the "fitness" of individuals from
different populations on a regular basis (every 4 month = \~8 gen). There are
three measures of fitness used, one of which is survival (following stress
treatment). This measure is designed to approximate how well a fly can survive
without the need for a full longevity assay (which may take months), and is thus
more suitable for a study of this size.

Flies from different populations (1-11,14) are left in a chill coma state for
25h (\~4ºC) - this is used as the stress treatment. Upon removal the flies are
returned to room temperature and left to recover. The number of surviving flies
per vial (of an original 10) are then counted after 5h (survival_5h) and 24h
(survival_24h).

## Reading in the data

```{r import_data}

cold_surv <- readxl::read_excel(
  path = "~/OneDrive - University College London/data/time_fitness/thermal/TIME_thermal_survival_data.xlsx", 
  col_names = TRUE
)

```

## Calculating survival fractions

```{r calculate_survival}

cold_surv %>% filter(survival_5h > start) # should be tibble with 0 rows 

cold_surv <- cold_surv %>% 
  mutate(
    # percentage survival 5 hours after removal from treatment
    per_5h = (survival_5h / start) * 100,
    # percentage survival 24 hours after removal from treatment
    per_24h = (survival_24h /start) * 100, 
    # survival percentage only including those that survived cold treatment
    per_drop = (survival_24h/survival_5h) * 100
  )

# set the order of genotypes
cold_surv$genotype <- factor(
  cold_surv$genotype, 
  levels = c("mM", "mX", "tX", "tT") 
)

# set the order of populations
cold_surv$pop <- factor(
  cold_surv$pop, 
  levels = c("11", "1", "2", "3","4", "5", "6", "7","8", "9", "10", "14") 
)
```

## Overview Plots

```{r general_boxplot, warning=FALSE, include=FALSE}

# listing what to plot
measures <- c("per_5h", "per_24h", "per_drop")

for (i in measures){  # for each variable in the measures vector:
  print(
    ggplot(
      data = cold_surv, 
      mapping = aes(
        x = genotype, 
        y = cold_surv[[i]], 
        colour = genotype
      )
    ) +
    geom_boxplot(
      outlier.shape = NA # do not plot outliers
    ) +
    geom_point(
      position = position_dodge2(0.25),
      alpha = 0.8
    ) + 
    scale_colour_manual(
      values = TIME_palette
    ) +
    ylim(
      c(0,100)
    ) +
    ylab(
      i
    ) +
    xlab(
      "Genotype"
    ) +
    facet_grid(
      cols = vars(timepoint)
    ) +
    TIME_theme
    ) # close print
}

```

## Population & Genotype Summaries

```{r surv_summaries}

surv_summary_pop <- cold_surv %>%
  group_by(
    genotype, pop, timepoint # group by timepoint and population
  ) %>%
  # calculate mean values for the two measurements
  reframe(
    mean_5h = mean( # mean 5h survival
      per_5h, 
      na.rm = TRUE
    ), 
    mean_24h = mean( # mean 24h survival
      per_24h, 
      na.rm = TRUE
    ), 
    mean_drop = mean( # mean survival only included those who survived treatment
      per_drop, 
      na.rm = TRUE
    ) , 
    # calculate the standard deviations for each population
    sd_5h = sd(
      per_5h, 
      na.rm = TRUE
    ), 
    sd_24h = sd(
      per_24h, 
      na.rm = TRUE
    ), 
    sd_drop = sd(
      per_drop, 
      na.rm = TRUE
    )
  )

# using the summary supplied above - calculate mean values per genotype (and timepoint)
surv_summary_genotype <- surv_summary_pop %>% 
  group_by(
    genotype, timepoint # group populations of the same genotype together
  ) %>%
  reframe(
    mean_5h = mean(
      mean_5h, 
      na.rm = TRUE
    ), 
    mean_24h = mean(
      mean_24h, 
      na.rm = TRUE
    ), 
    mean_drop = mean(
      mean_drop, 
      na.rm = TRUE
    ),
    sd_5h = sd(
      sd_5h, 
      na.rm = TRUE
    ), 
    sd_24h = sd(
      sd_24h, 
      na.rm = TRUE
    ), 
    sd_drop = sd(
      sd_drop, 
      na.rm = TRUE
    ), 
    N = n()
  )

```

## Normalising by mM

As there is variability in the parental lines (unexpected, but likely due to the
effects of living in the cages) an internal control can be used to compare
relative differences.

```{r normalising_by_mM}

# create small dataframe with mean values for mM and tT
mMtT_surv <- surv_summary_pop %>%
  group_by(
    timepoint
  ) %>%
  filter(
    genotype %in% c("mM", "tT") # select the internal controls 
  ) %>%
  group_by(
    timepoint
  )%>%
  reframe(
    mean_drop = mean(mean_drop) # calculate an average of these
  ) %>%
  select(timepoint, mean_drop)

# join this to the full data 
surv_normalised <- left_join(
  x = surv_summary_pop, 
  y = mMtT_surv, 
  by = "timepoint"
)

# normalise the values 
surv_normalised <- surv_normalised %>%
  reframe(
    genotype = genotype, 
    pop = pop,
    timepoint = timepoint, 
    surv_norm = mean_drop.x/mean_drop.y,
    upper = (mean_drop.x + sd_drop)/mean_drop.y,
    lower = (mean_drop.x - sd_drop)/mean_drop.y
  )

# calculate averages by genotype 
surv_normalised_gt <- surv_normalised %>%
  group_by(
    timepoint, genotype
  ) %>%
  reframe(
    surv_norm_gt = mean(surv_norm),
    sd_norm = sd(surv_norm),
    upper = (surv_norm_gt + sd_norm),
    lower = (surv_norm_gt - sd_norm)
  )

```

```{r normalised_plots}

# where y = y_value, error_bar_value, y_lower_limit, y_upper_limit
#y <- c("surv_norm", "upper", "lower", "pop")
y <- c("surv_norm_gt", "upper", "lower", "genotype")

ggplot(
  data = surv_normalised_gt, 
  mapping = aes(
    x = .data[[y[4]]], 
    y = .data[[y[1]]], 
    colour = genotype
  )
) +
  geom_point(
    size = 4
  ) + 
  geom_errorbar(
    data = surv_normalised_gt,
    na.rm = TRUE,
    mapping = aes(
      ymin = .data[[y[2]]], 
      ymax = .data[[y[3]]]
    )
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(timepoint)
  ) +
  TIME_theme
```

```{r line_plots}

y <- c("surv_norm_gt", "sd_norm")

ggplot(
  data = surv_normalised_gt, 
  mapping = aes(
    x = timepoint, 
    y = .data[[y[1]]], 
    colour = as.factor(genotype), 
    #linetype = as.factor(genotype)
  )
) +
  geom_ribbon(
    mapping = aes(
      ymin = .data[[y[1]]] - .data[[y[2]]], 
      ymax = .data[[y[1]]] + .data[[y[2]]], 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    linewidth = 2
  ) + 
  geom_point(
    size = 4
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
   scale_fill_manual(
    values = TIME_palette
  ) +
   #facet_grid(
  #  cols = vars(genotype)
  #)  +
  TIME_theme

```

## Non-normalised Graphs (NOT RECOMMENDED)

```{r plotting_summaries_pop, warning=FALSE}

y <- c("mean_5h", "sd_5h")
y <- c("mean_24h", "sd_5h")
y <- c("mean_drop", "sd_drop")

ggplot(
  data = surv_summary_pop, 
  mapping = aes(
    x = as.factor(pop), 
    y = .data[[y[1]]], 
    colour = genotype
  )
) +
  geom_errorbar(
    #data = surv_summary_pop,
    mapping = aes(
      ymin = .data[[y[1]]] - .data[[y[2]]], 
      ymax = .data[[y[1]]] + .data[[y[2]]]
    )
  ) +
    geom_point(
      size = 5
  ) + 
  ylab(
    y[1]
  ) +
  xlab(
    "Population"
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(timepoint)
  ) + 
  theme_light()
```

```{r plotting_summaries_genotype}

# where y = y_value, error_bar_value, y_lower_limit, y_upper_limit
#y <- c("mean_24h", "sd_24h", 0, 100)
#y <- c("mean_5h", "sd_5h", 90, 100)
y <- c("mean_drop", "sd_drop", 0, 100)

ggplot(
  data = surv_summary_genotype, 
  mapping = aes(
    x = genotype, 
    y = .data[[y[1]]], 
    colour = genotype
  )
) +
  geom_point(
    size = 4
  ) + 
  geom_errorbar(
    data = surv_summary_genotype,
    na.rm = TRUE,
    mapping = aes(
      ymin = .data[[y[1]]] - .data[[y[2]]], 
      ymax = .data[[y[1]]] + .data[[y[2]]]
    )
  ) +
  ylim(
    c(
      as.numeric(y[3]), 
      as.numeric(y[4])
    )
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(timepoint)
  ) +
  TIME_theme
```
