---
title: "DAMs_stats"
author: "FiG-T"
date: "`r Sys.Date()`"
output: 
html_document: 
  keep_md: yes
    toc: yes
  toc: TRUE
  editor_options:
  markdown:
    wrap: 80
  chunk_output_type: inline
---

# Libraries & Setup

## Required Libraries

```{r libraries}

library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)

```

## Theme setup

```{r setup}

# set the palette
TIME_palette <- c(
  mX = "orchid1", 
  tX = "darkorchid", 
  mM = "deepskyblue2", 
  tT = "firebrick3"
)

col2rgb(TIME_palette)

TIME_theme <- theme_light(
) +
  theme(
    strip.background = element_rect(
      fill = "snow4"
    )
  )
```

# Formatting & Analysis of Drosophila Activity Data

## Introduction

Drosophila Activity Monitors (a.k.a DAMs) (TriKinetics) are a valuable tool for assessing the patterns of movement in fly studies. Derived metrics (peak/average activity, number of sleeps etc. etc) are frequently used as a proxy for "fitness". One advantage of the DAMs is that each monitor can simultaneously measure up to 32 flies, and multiple monitors can be used at once. DAMs are thus suitable for scaling for use in high-throughput studies.

**The following script is designed for use for formatting and analysing extracted data (from raw output files).**

If the data is still in the raw data (MonitorX.txt) format, please see 'import_DAMS.Rmd' for details on how to convert this to a usable structure.

## Importing Data

```{r import_data}

# load in data
dams_table <- readxl::read_excel(
  path = "~/OneDrive - University College London/data/time_fitness/activity/time_activity_data.xlsx"
)
```

## Formatting data

So that the downstream analysis and plotting will work correctly the data must be formatted.

```{r data_formatting}

# setting genotypes and populations to factors
# set the order of genotypes
dams_table$genotype <- factor(
  dams_table$genotype, 
  levels = c("mM", "mX", "tX", "tT") 
)

# set the order of populations
dams_table$pop <- factor(
  dams_table$pop, 
  levels = c("11", "1", "2", "3","4", "5", "6", "7","8", "9", "10", "14") 
)


```

### Selecting experimental days

```{r select_day_2}

# create a vector of all dates in the dataset (already prefiltered in the input function)
exp_dates <- unique(dams_table$date)

# extract the experimental dates
day2_activity <- dams_table %>%
  filter(
    date %in% exp_dates[c(3,8,12)]   # change this to desired date
  )

# check that the correct dates are included
unique(day2_activity$date)

# check that all monitors were working on the selected days
unique(day2_activity$status) == 1

# pivot into a wider format
day2_activity <- day2_activity %>%
  tidyr::pivot_longer(
    cols = matches("chamber.*"), 
    names_to = "chamber", 
    values_to = "counts"
  )

# convert monitor to a numeric variable
day2_activity$chamber <- as.numeric(
  str_remove_all(
    day2_activity$chamber, 
    pattern = "[:alpha:]|[:punct:]"
  )
)

```

## Total Activity

The absence of any counts (or presence of a very limited number) across the day may be indicative of an injured, dead, or trapped fly.

```{r sum_activity_chamber}

# average activity by monitor across the whole day
sum_activity_chamber <- day2_activity %>%
  group_by(
    bloc, monitor, chamber
  ) %>%
  reframe(
    sum_counts = sum(counts)
  )

# plot the total activity counts
ggplot2::ggplot(
  data = sum_activity_chamber, 
  mapping = aes(
    x = chamber, 
    y = sum_counts, 
    colour = as.factor(monitor)
  )
) + 
  geom_point(
) + 
  facet_wrap(
    ~monitor
  ) +
  ylim(
    c(0,20)
  )
```

```{r filter_total}

# merge total sum with full time data
day2_activity <- left_join(
  day2_activity, 
  sum_activity_chamber
)

# set threshold
count_threshold <- 150

# remove chambers that do not meet the threshold number
day2_activity <- day2_activity %>%
  filter(
    sum_counts >= count_threshold
  ) %>%
  select(
    -sum_counts
  )

```

## Average Activity

Average activity can be calculated in two ways:

1.  Rolling averages/sums (sliding 30 min windows)
2.  Block averages/sums (non-overlapping 30 min windows)

### Activity by monitor

```{r rolling_averages}

# duplicate data
day2_activity_rollmean <- day2_activity 

# convert time to 'time' data class 
day2_activity_rollmean$time <- hms::as_hms(day2_activity_rollmean$time)

# calculate rolling averages
day2_activity_rollmean <- day2_activity_rollmean %>%
  group_by(
     bloc, date, monitor
  ) %>%
  mutate(
    mean_count_60 = zoo::rollmean( # mean per hour window
      counts, 
      k = (60*60), 
      fill = NA, 
      align = "center", 
      na.rm = FALSE
    )
  ) %>%
  mutate(
    sum_count_30 = zoo::rollsum( # sum per 30 min window
      counts, 
      k = (60*30), 
      fill = NA, 
      align = "center"
    )
  ) %>%
  mutate(
    mean_count_30 = zoo::rollmean( # mean count per 30 min window
      counts, 
      k = (30*60), 
      fill = NA, 
      align = "center"
    )
  ) %>%
  mutate(
    mean_count_15 = zoo::rollmean( # mean count per 30 min window
      counts, 
      k = (60*15), 
      fill = NA, 
      align = "center"
    )
  )
```

```{r block_averages}

# duplicate dataset
day2_activity_mean <- day2_activity 

# format time into the correct time/date format 
day2_activity_mean$time <- as.POSIXct(
  day2_activity_mean$time, 
  format = "%H:%M:%S"
)

# add in time intervals
day2_activity_mean <- day2_activity_mean %>%
  dplyr::mutate(
    time_interval = cut(
      time, 
      breaks = "30 min"
    )
  ) %>%
  group_by(
    time_interval, bloc, monitor
  ) %>%
  reframe(
    sum_count_30 = sum(counts),
    mean_count_30 = mean(counts),
    time_interval = time_interval, 
    genotype = genotype, 
    pop = pop, 
    light = light,
  )  %>%
  distinct()

# convert time interval to character (from numeric)
day2_activity_mean$time_interval <- as.character(
    day2_activity_mean$time_interval
    )

# reconvert to date
day2_activity_mean$time_interval <- ymd_hms(
    day2_activity_mean$time_interval
    )

# reconvert to correct format
day2_activity_mean$time_interval <- format(
    day2_activity_mean$time_interval, 
    "%H:%M"
    )

# remove rows where no time data is present
day2_activity_mean <- day2_activity_mean %>%
  filter(! is.na(time_interval))

# reformat time (again)
day2_activity_mean$time_interval <- hm(
    day2_activity_mean$time_interval
)

```

### Activity by genotype

```{r activity_by_genotype}

# grouped mean
day2_activity_gt <- day2_activity_mean %>%
  group_by(
    time_interval, bloc, genotype 
  ) %>%
  reframe(
    genotype = genotype,
    light = light,
    mean_window = mean(mean_count_30), 
    sd_gt = sd(mean_count_30)
  ) %>%
  distinct()


# rolling mean 
day2_activity_gt_roll <- day2_activity_rollmean %>%
  group_by(
    time, bloc, genotype 
  ) %>%
  reframe(
    genotype = genotype, 
    light = light,
    mean_window = mean(mean_count_30), 
    sd_gt = sd(mean_count_30)
  ) %>%
  distinct()

```

## Plotting activity

### Population Plots

```{r monitor_plots, include=FALSE}

# for grouped mean
ggplot2::ggplot(
  data = day2_activity_mean, 
  mapping = aes(
    x = time_interval, 
    y = mean_count_30, 
    linetype = pop
  )
) + geom_line(
  mapping = aes(
    colour = genotype
  )
) +
  ylab(
    "Blocked Mean of Counts (30 min windows)"
  ) +
  scale_colour_manual(
    values = TIME_palette
  )

# for rolling mean
ggplot2::ggplot(
  data = day2_activity_rollmean %>%
    filter(pop == "1"), 
  mapping = aes(
    x = time, 
    y = mean_count_30, 
    linetype = pop
  )
) + geom_line(
  mapping = aes(
    colour = genotype
  )
) +
  ylab(
    "Rolling Mean of Counts (30 min windows)"
  ) +
  facet_grid(
    rows = vars(bloc)
  ) +
  scale_colour_manual(
    values = TIME_palette
  )
```

### Plots by genotype

```{r plotting_gt_activity, warning=FALSE}

# for block means
ggplot2::ggplot(
  data = day2_activity_gt, 
  mapping = aes(
    x = time_interval, 
    y = mean_window, 
    colour = genotype
  )
) + 
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    mapping = aes(
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    mapping = aes(
      colour = genotype
    ), 
    #position = position_dodge2(50)
) +
  geom_point(
    mapping = aes(
      y = mean_window, 
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    ) 
  ) +
  facet_grid(
    rows = vars(bloc)
  ) + 
  TIME_theme


# for rolling means
ggplot2::ggplot(
) +
  geom_rect(
    data = light_timings,
    mapping = aes(
      ymin = -Inf,
      ymax = Inf, 
      xmin = xstart,
      xmax = xend,
      alpha = as.factor(light)
    ), 
    show.legend = FALSE
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.1)
  ) +
    facet_grid(
    rows = vars(bloc)
  ) + 
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    data = day2_activity_gt_roll,
    mapping = aes(
      x = time,
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    )
) +
  geom_point(
    data = day2_activity_gt_roll , 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    )
  ) +
  TIME_theme
```

## Activity data stats

```{r labels}


light_labels <- labeller(
  light = c(
    "0" = "Night", 
    "1" = "Day"
  )
)
```

## Light timings

\-\-\-\-\-\-\-\-\-\-\-\-\-- Need to update --------------------

```{r light_timings}

start <- day2_activity_gt_roll %>% # calculated below
  slice_min(time) %>%
  select(time) 

light_on <- day2_activity_gt_roll %>%
  filter(light == 1) %>%
  slice_min(time) %>%
  select(time) 

light_off <- day2_activity_gt_roll %>%
  filter(light == 1) %>%
  slice_max(time) %>%
  select(time) 

end <- day2_activity_gt_roll %>%
  slice_max(time) %>%
  select(time) 

start <- unique(start$time)
light_on <- unique(light_on$time)
light_off <- unique(light_off$time)
end <- unique(end$time)

light_timings <- tibble(
  xstart = c(start, light_on, light_off), 
  xend = c(light_on, light_off , end), 
  light = c(0,1,0)
)

```

## Activity plots

-   by monitor

-   by genotype

```{r monitor_plots_depricated}

# for grouped mean
ggplot2::ggplot(
  data = day2_monitor_mean, 
  mapping = aes(
    x = time_interval, 
    y = mean_window, 
    linetype = as.factor(pop)
  )
) + geom_line(
  mapping = aes(
    colour = as.factor(pop)
  )
) +
  ylab(
    "Blocked Mean of Counts (30 min windows)"
  ) #+
  scale_colour_manual(
    values = TIME_palette
  )

# for rolling mean
ggplot2::ggplot(
  data = day2_monitor_rollmean, 
  mapping = aes(
    x = time, 
    y = mean_window, 
    linetype = as.factor(pop)
  )
) + geom_line(
  mapping = aes(
    colour = genotype
  )
) +
  ylab(
    "Rolling Mean of Counts (30 min windows)"
  ) +
  scale_colour_manual(
    values = TIME_palette
  )

```

```{r plotting_gt_activity_depricated}

# for block means
ggplot2::ggplot(
  data = day2_activity_gt, 
  mapping = aes(
    x = time_interval, 
    y = mean_window, 
    colour = genotype
  )
) + 
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    mapping = aes(
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    mapping = aes(
      colour = genotype
    ), 
    #position = position_dodge2(50)
) +
  geom_point(
    mapping = aes(
      y = mean_window, 
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    )
  ) + theme_minimal()


# for rolling means
ggplot2::ggplot(
) +
  geom_rect(
    data = light_timings,
    mapping = aes(
      ymin = -Inf,
      ymax = Inf, 
      xmin = xstart,
      xmax = xend,
      alpha = as.factor(light)
    ), 
    show.legend = FALSE
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.1)
  ) +
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    data = day2_activity_gt_roll,
    mapping = aes(
      x = time,
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    )
) +
  geom_point(
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    )
  ) 
  facet_grid(
    rows = vars(bloc)
  )

```

## Grouping Day & Night

As a single metric is often easier to interpret than a full activity plot...

```{r day_night_per_pop}

# for block means
diurnal_activity <- day2_activity_mean %>%
  group_by(
    light, bloc, genotype, pop 
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    light = light,
    mean_window = mean(sum_count_30), 
    sd_pop = sd(sum_count_30)
  ) %>%
  distinct()

# for rolling averages
diurnal_activity_roll <- day2_activity_rollmean %>%
   group_by(
    light, bloc, genotype, pop
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    light = light,
    mean_window = mean(mean_count_60, na.rm = TRUE), 
    sd_pop = sd(mean_count_60, na.rm = TRUE)
  ) %>%
  distinct()

# for rolling averages
diurnal_activity_roll_chamber <- day2_activity_rollmean %>%
   group_by(
    light, bloc, genotype, pop, chamber
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    light = light,
    mean_window = mean(mean_count_60, na.rm = TRUE), 
    sd_pop = sd(mean_count_60, na.rm = TRUE)
  ) %>%
  distinct()

  
```

```{r plot_day_vs_night}

diurnal_activity_pop <- list(diurnal_activity, diurnal_activity_roll)

for (i in c(1,2)) {
  
  print(
  ggplot2::ggplot(
  data = diurnal_activity_pop[[i]], 
  mapping = aes(
    x = pop, 
    y = mean_window, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = mean_window - sd_pop, 
      ymax = mean_window + sd_pop
    ), 
    width = 0.5
  ) +
  geom_point(
    size = 5
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  ylab("Mean activity")+
  facet_grid(
    cols = vars(bloc), 
    rows = vars(light),
    labeller = light_labels
  ) +
  theme_light()

  ) # close print
}

```

### Grouping by genotype

```{r diurnal_by_gt}

# for block means
diurnal_activity_gt <- diurnal_activity %>%
  group_by(
    light, bloc, genotype 
  ) %>%
  reframe(
    genotype = genotype,
    light = light,
    mean_activity = mean(mean_window), 
    sd_gt = sd(mean_window)
  ) %>%
  distinct()

# for rolling averages
diurnal_activity_roll_gt <- diurnal_activity_roll %>%
   group_by(
    light, bloc, genotype
  ) %>%
  reframe(
    genotype = genotype,
    light = light,
    mean_activity = mean(mean_window, na.rm = TRUE), 
    sd_gt = sd(mean_window, na.rm = TRUE)
  ) %>%
  distinct()

```

```{r diurnal_plots_gt}

diurnal_activity_gt <- list(diurnal_activity_gt, diurnal_activity_roll_gt)

for (i in c(1,2)) {
  
  print(
  ggplot2::ggplot(
  data = diurnal_activity_gt[[i]], 
  mapping = aes(
    x = genotype, 
    y = mean_activity, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = mean_activity - sd_gt, 
      ymax = mean_activity + sd_gt
    ), 
    width = 0.5
  ) +
  geom_point(
    size = 5
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(bloc), 
    rows = vars(light),
    labeller = light_labels
  ) +
  theme_light()

  ) # close print
}

```

### Looking at dawn and dusk

```{r pop_peak_activity}

peak_activity <- day2_activity %>%
  filter(
      time == '10 hours' | time == '21.5 hours'
  ) %>%
  filter(
    light == 0
  ) %>%
   group_by(
    time_interval, bloc, genotype, pop
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    peak_activity = mean(sum_count_30, na.rm = TRUE), 
    sd_pop = sd(sum_count_30, na.rm = TRUE)
  ) %>%
  distinct()

# for rolling averages
peak_activity <- day2_activity
peak_activity$time <- hms::as_hms(peak_activity$time)

peak_activity <- peak_activity %>%
  group_by(
     bloc, date, monitor, chamber
  ) %>%
  mutate(
    sum_count_30 = zoo::rollsum( # sum per 30 min window
      counts, 
      k = (30), 
      fill = NA, 
      align = "center"
    )
  )

peak_activity <- peak_activity %>%
  filter(
      time == (60*60*10) | time == (60*60*21.5)
  ) %>%
   group_by(
    time, bloc, genotype, pop
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    peak_activity = mean(sum_count_30, na.rm = TRUE), 
    sd_pop = sd(sum_count_30, na.rm = TRUE)
  ) %>%
  distinct()

```

```{r}
# per monitor
ggplot(
  data = peak_activity, 
  mapping = aes(
    x = pop, 
    y = peak_activity, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = peak_activity - sd_pop, 
      ymax = peak_activity + sd_pop
    ), 
    width = 0.2, 
    position = position_dodge(0.4)
  ) +
  geom_point(
    position = position_dodge(0.4)
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(time), 
    rows = vars(bloc)
  ) +
  TIME_theme

```

```{r dawn_dusk_sum}

# for grouped mean
peak_activity_gt <- day2_activity_mean %>%
  filter(
      time_interval == '10 hours' | time_interval == '21.5 hours'
  ) %>%
  group_by(
    time_interval, bloc, genotype
  ) %>%
  reframe(
    genotype = genotype,
    time = time_interval,
    mean_window = mean(mean_count_30), 
    sd_gt = sd(mean_count_30)
  ) %>%
  distinct()

# for rolling averages
peak_activity_gt_roll <- peak_activity %>%
 # filter(
#      time == '10 hours' | time == '21.5 hours'
#  ) %>%
   group_by(
    time, bloc, genotype
  ) %>%
  reframe(
    genotype = genotype,
    peak_activity_mean = mean(peak_activity, na.rm = TRUE), 
    sd_gt = sd(peak_activity, na.rm = TRUE)
  ) %>%
  distinct()



```

```{r dawndusk_plot}

# per monitor
ggplot(
  data = peak_activity_gt, 
  mapping = aes(
    x = as.factor(pop), 
    y = monitor_mean, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = monitor_mean - sd_gt, 
      ymax = monitor_mean + sd_gt
    ), 
    width = 0.2, 
    position = position_dodge(0.4)
  ) +
  geom_point(
    position = position_dodge(0.4)
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(as.character(time_interval))
  ) +
  theme_light()

# per genotype
ggplot(
  data = peak_activity_gt_roll, 
  mapping = aes(
    x = genotype, 
    y = peak_activity_mean, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = peak_activity_mean - sd_gt, 
      ymax = peak_activity_mean + sd_gt
    ), 
    width = 0.2, 
    position = position_dodge(0.4)
  ) +
  geom_point(
    position = position_dodge(0.4), 
    size = 3
  ) + 
  ylab(
    "Mean Activity"
  ) +
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(time), 
    rows = vars(bloc)
  ) +
  theme_light()
  
  
  

```

```{r dawn_dusk_lm}

head(diurnal_activity_roll_chamber)

#mod_data <- dawn_activity_chamber %>%
#  filter(time_interval > "10 hours" )

mod_data <- diurnal_activity_roll_chamber %>%
  filter(light == 1)

for (i in c("bloc","monitor" ,"genotype", "pop", "chamber")) {
  mod_data[[i]] <- as.factor(mod_data[[i]])
}

mod_data[["chamber"]] <- as.factor(mod_data[["chamber"]])
str(mod_data)

activity_mod <- lme4::lmer(
  formula = mean_window ~   bloc + genotype + pop + (1|chamber),
  data = mod_data
)
mod_aov <- car::Anova(activity_mod)
mod_aov

mod_emm <- emmeans::emmeans(
  activity_mod, 
  pairwise ~ genotype, 
  adjust = "tukey"
)
mod_emm


###-----------------------------------------

```
