---
title: "DAMs_stats"
author: "FiG-T"
date: "`r Sys.Date()`"
output: 
html_document: 
  keep_md: yes
    toc: yes
  toc: TRUE
  editor_options:
  markdown:
    wrap: 80
  chunk_output_type: inline
---

# Libraries & Setup

## Required Libraries

```{r libraries}

library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)

```

## Theme setup

```{r time_colours}

source("https://raw.githubusercontent.com/FiG-T/scripts/refs/heads/main/R/time_r/time_palettes_themes.R")
```

# Formatting & Analysis of Drosophila Activity Data

## Introduction

Drosophila Activity Monitors (a.k.a DAMs) (TriKinetics) are a valuable tool for assessing the patterns of movement in fly studies. Derived metrics (peak/average activity, number of sleeps etc. etc) are frequently used as a proxy for "fitness". One advantage of the DAMs is that each monitor can simultaneously measure up to 32 flies, and multiple monitors can be used at once. DAMs are thus suitable for scaling for use in high-throughput studies.

**The following script is designed for use for formatting and analysing extracted data (from raw output files).**

If the data is still in the raw data (MonitorX.txt) format, please see 'import_DAMS.Rmd' for details on how to convert this to a usable structure.

## Importing Data

```{r import_data}

# load in data
dams_table <- readxl::read_excel(
  path = "~/OneDrive - University College London/data/time_fitness/activity/time_activity_data_bloc4.xlsx"
)
```

## Formatting data

So that the downstream analysis and plotting will work correctly the data must be formatted.

```{r data_formatting}

# setting genotypes and populations to factors
# set the order of genotypes
dams_table$genotype <- factor(
  dams_table$genotype, 
  levels = c("mM", "mX", "tX", "tT") 
)

# set the order of populations
dams_table$pop <- factor(
  dams_table$pop, 
  levels = c("11", "1", "2", "3","4", "5", "6", "7","8", "9", "10", "14") 
)


```

### Selecting experimental days

```{r select_day_2}

# create a vector of all dates in the dataset (already prefiltered in the input function)
exp_dates <- unique(dams_table$date)

# extract the experimental dates
day2_activity <- dams_table %>%
  filter(
    date %in% exp_dates[c(3,8,12,17)]   # change this to desired date
  )

# check that the correct dates are included
unique(day2_activity$date)

# check that all monitors were working on the selected days
unique(day2_activity$status) == 1

# pivot into a wider format
day2_activity <- day2_activity %>%
  tidyr::pivot_longer(
    cols = matches("chamber.*"), 
    names_to = "chamber", 
    values_to = "counts"
  )

# convert monitor to a numeric variable
day2_activity$chamber <- as.numeric(
  str_remove_all(
    day2_activity$chamber, 
    pattern = "[:alpha:]|[:punct:]"
  )
)

```

## Total Activity

The absence of any counts (or presence of a very limited number) across the day may be indicative of an injured, dead, or trapped fly.

```{r sum_activity_chamber}

# sum activity by monitor across the whole day
sum_activity_chamber <- day2_activity %>%
  group_by(
    bloc, monitor, chamber
  ) %>%
  reframe(
    sum_counts = sum(counts)
  )

# plot the total activity counts
ggplot2::ggplot(
  data = sum_activity_chamber, 
  mapping = aes(
    x = chamber, 
    y = sum_counts, 
    colour = as.factor(monitor)
  )
) + 
  geom_point(
) + 
  facet_wrap(
    ~monitor
  ) +
  ylim(
    c(0,20)
  )
```

```{r filter_total}

# merge total sum with full time data
day2_activity <- left_join(
  day2_activity, 
  sum_activity_chamber
)

# set threshold
count_threshold <- 150

# remove chambers that do not meet the threshold number
day2_activity <- day2_activity %>%
  filter(
    sum_counts >= count_threshold
  ) %>%
  select(
    -sum_counts # then remove this column
  )

```

## Average Activity

Average activity can be calculated in two ways:

1.  Rolling averages/sums (sliding 30 min windows)
2.  Block averages/sums (non-overlapping 30 min windows)

### Activity by monitor

```{r rolling_averages}

# duplicate data (to backup against any future errors)
day2_activity_rollmean <- day2_activity 

# convert time to 'time' data class 
day2_activity_rollmean$time <- hms::as_hms(day2_activity_rollmean$time)

# calculate rolling averages
day2_activity_rollmean <- day2_activity_rollmean %>%
  group_by(
     bloc, date, monitor
  ) %>%
  mutate(
    mean_count_60 = zoo::rollmean( # mean per hour window
      counts, 
      k = (60*60), # time in seconds 
      fill = NA, 
      align = "center", 
      na.rm = FALSE
    )
  ) %>%
  mutate(
    sum_count_30 = zoo::rollsum( # sum per 30 min window
      counts, 
      k = (60*30), # time in seconds
      fill = NA, 
      align = "center"
    )
  ) %>%
  mutate(
    mean_count_30 = zoo::rollmean( # mean count per 30 min window
      counts, 
      k = (30*30), 
      fill = NA, 
      align = "center"
    )
  ) %>%
  mutate(
    mean_count_15 = zoo::rollmean( # mean count per 30 min window
      counts, 
      k = (60*15), 
      fill = NA, 
      align = "center"
    )
  )
```

If you wish to use the "block approach" then the following code can be used. This method will result in less smooth averages and as such is not recommended.

```{r block_averages, include = FALSE}

# duplicate dataset
day2_activity_mean <- day2_activity 

# format time into the correct time/date format 
day2_activity_mean$time <- as.POSIXct(
  day2_activity_mean$time, 
  format = "%H:%M:%S"
)

# add in time intervals
day2_activity_mean <- day2_activity_mean %>%
  dplyr::mutate(
    time_interval = cut(
      time, 
      breaks = "30 min"
    )
  ) %>%
  group_by(
    time_interval, bloc, monitor
  ) %>%
  reframe(
    sum_count_30 = sum(counts),
    mean_count_30 = mean(counts),
    time_interval = time_interval, 
    genotype = genotype, 
    pop = pop, 
    light = light,
  )  %>%
  distinct()

# convert time interval to character (from numeric)
day2_activity_mean$time_interval <- as.character(
    day2_activity_mean$time_interval
    )

# reconvert to date
day2_activity_mean$time_interval <- ymd_hms(
    day2_activity_mean$time_interval
    )

# reconvert to correct format
day2_activity_mean$time_interval <- format(
    day2_activity_mean$time_interval, 
    "%H:%M"
    )

# remove rows where no time data is present
day2_activity_mean <- day2_activity_mean %>%
  filter(! is.na(time_interval))

# reformat time (again)
day2_activity_mean$time_interval <- hm(
    day2_activity_mean$time_interval
)

```

### Activity by genotype

```{r activity_by_genotype}

# rolling mean 
day2_activity_gt_roll <- day2_activity_rollmean %>%
  group_by(
    time, bloc, genotype 
  ) %>%
  reframe(
    genotype = genotype, 
    light = light,
    mean_window = mean(mean_count_30), 
    sd_gt = sd(mean_count_30)
  ) %>%
  distinct()

# grouped mean
#day2_activity_gt <- day2_activity_mean %>%
#  group_by(
#    time_interval, bloc, genotype 
#  ) %>%
#  reframe(
#    genotype = genotype,
#    light = light,
#    mean_window = mean(mean_count_30), 
#    sd_gt = sd(mean_count_30)
#  ) %>%
#  distinct()

```

## Light timings

```{r labels}


light_labels <- labeller(
  light = c(
    "0" = "Night", 
    "1" = "Day"
  )
)
```

```{r light_timings}

# select the start time for the recording 
start <- day2_activity_gt_roll %>% # calculated below
  slice_min(time) %>%
  select(time) 

# determine when the light comes on
light_on <- day2_activity_gt_roll %>%
  filter(bloc != 4) %>% # remove bloc 4 due to faulty lighting
  filter(light == 1) %>%
  slice_min(time) %>% # when does the light turn on?
  select(time) 

# determine when the light turns off
light_off <- day2_activity_gt_roll %>%
  filter(bloc != 4) %>% # still exclude bloc 4
  filter(light == 1) %>%
  slice_max(time) %>%
  select(time) 

# determine the end of the recording window
end <- day2_activity_gt_roll %>%
  slice_max(time) %>%
  select(time) 

# convert from tibble to character
start <- unique(start$time)
light_on <- unique(light_on$time)
light_off <- unique(light_off$time)
end <- unique(end$time)

# combine together 
light_timings <- tibble(
  xstart = c(start, light_on, light_off), 
  xend = c(light_on, light_off , end), 
  light = c(0,1,0)
)

```

## Plotting activity

### Population Plots

```{r monitor_plots, include=FALSE}

# for rolling mean
ggplot2::ggplot(
  data = day2_activity_rollmean %>%
    filter(pop == "1"), 
  mapping = aes(
    x = time, 
    y = mean_count_30, 
    linetype = pop
  )
) + geom_line(
  mapping = aes(
    colour = genotype
  )
) +
  ylab(
    "Rolling Mean of Counts (30 min windows)"
  ) +
  facet_grid(
    rows = vars(bloc)
  ) +
  scale_colour_manual(
    values = TIME_palette
  )
```

This graph show the average for a sample population (here mX_1), there is a noticeable issue for bloc 4. This is highly likely due to errors in the Temperature Cabinet (a different one was used for thus timepoint).

### Plots by genotype

```{r plotting_gt_activity, warning=FALSE}

# for rolling means
ggplot2::ggplot(
) +
  geom_rect( # set the background colours for light and dark 
    data = light_timings,
    mapping = aes(
      ymin = -Inf, # cover the whole of the yaxis 
      ymax = Inf, 
      xmin = xstart,
      xmax = xend,
      alpha = as.factor(light)
    ), 
    show.legend = FALSE
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.1)
  ) +
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    data = day2_activity_gt_roll, # using the activity summary dataset 
    mapping = aes(
      x = time,
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line( # overlay the lines 
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    )
) +
  geom_point( # add in the timepoint dots 
    data = day2_activity_gt_roll , 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  facet_grid(
    rows = vars(bloc)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  #breaks = c(
  #    hours
  # )
  #) +
  TIME_theme
```

## Grouping Day & Night

As a single metric is often easier to interpret than a full activity plot...

```{r day_night_per_pop}

# for block means
#diurnal_activity <- day2_activity_mean %>%
#  group_by(
#    light, bloc, genotype, pop 
#  ) %>%
#  reframe(
#    genotype = genotype,
#    pop = pop,
#    light = light,
#    mean_window = mean(sum_count_30), 
#    sd_pop = sd(sum_count_30)
#  ) %>%
#  distinct()

# for rolling averages
diurnal_activity_roll <- day2_activity_rollmean %>%
   group_by(
    light, bloc, genotype, pop  # use the same groupings as earlier but split by light 
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    light = light,
    mean_window = mean(mean_count_60, na.rm = TRUE), 
    sd_pop = sd(mean_count_60, na.rm = TRUE)
  ) %>%
  distinct()

# for rolling averages
diurnal_activity_roll_chamber <- day2_activity_rollmean %>%
   group_by(
    light, bloc, genotype, pop, chamber # extra chamber grouping factor 
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    light = light,
    mean_window = mean(mean_count_60, na.rm = TRUE), 
    sd_pop = sd(mean_count_60, na.rm = TRUE)
  ) %>%
  distinct()

  
```

```{r plot_day_vs_night}

# initiate a list (extra dataframes can be added if required)
diurnal_activity_pop <- list(diurnal_activity_roll)

# loop through dataframes in list
for (i in c(1)) {
  
  print(
  ggplot2::ggplot(
    data = diurnal_activity_pop[[i]], 
    mapping = aes(
      x = pop, 
      y = mean_window, 
      colour = genotype
    )
  ) +
  geom_errorbar(
    mapping = aes(
      ymin = mean_window - sd_pop, 
      ymax = mean_window + sd_pop
    ), 
    width = 0.5
  ) +
  geom_point(
    size = 5
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  ylab("Mean activity")+
  facet_grid(
    cols = vars(bloc), 
    rows = vars(light),
    labeller = light_labels
  ) +
  theme_light()

  ) # close print
}

```

### Grouping by genotype

```{r diurnal_by_gt}

# for block means
#diurnal_activity_gt <- diurnal_activity %>%
#  group_by(
#    light, bloc, genotype 
#  ) %>%
#  reframe(
#    genotype = genotype,
#    light = light,
#    mean_activity = mean(mean_window), 
#    sd_gt = sd(mean_window)
#  ) %>%
#  distinct()

# for rolling averages
diurnal_activity_roll_gt <- diurnal_activity_roll %>%
   group_by(
    light, bloc, genotype
  ) %>%
  reframe(
    genotype = genotype,
    light = light,
    mean_activity = mean(
      mean_window, 
      na.rm = TRUE
    ), 
    sd_gt = sd(
      mean_window, 
      na.rm = TRUE
    )
  ) %>%
  distinct()

```

```{r diurnal_plots_gt}

diurnal_activity_gt <- list(diurnal_activity_roll_gt)

for (i in c(1)) {
  
  print(
  ggplot2::ggplot(
  data = diurnal_activity_gt[[i]], 
  mapping = aes(
    x = genotype, 
    y = mean_activity, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = mean_activity - sd_gt, 
      ymax = mean_activity + sd_gt
    ), 
    width = 0.5
  ) +
  geom_point(
    size = 5
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(bloc), 
    rows = vars(light),
    labeller = light_labels
  ) +
  theme_light()

  ) # close print
}

```
Here the failure of the lights to turn off in Bloc 4 is clear... 

### Looking at dawn and dusk

```{r pop_peak_activity}

# for rolling averages
peak_activity <- day2_activity
peak_activity$time <- hms::as_hms(peak_activity$time)

# recalculate the counts per 30 min window
peak_activity <- peak_activity %>%
  group_by(
     bloc, date, monitor, chamber
  ) %>%
  mutate(
    sum_count_30 = zoo::rollsum( # sum per 30 min window
      counts, 
      k = (30), 
      fill = NA, 
      align = "center"
    )
  )

# reduce this dataset to filter for the peak times
peak_activity <- peak_activity %>%
  filter(
      time == (60*60*10) | time == (60*60*21.5) # select 10AM and 9:30PM
  ) %>%
   group_by(
    time, bloc, genotype, pop
  ) %>%
  reframe(
    genotype = genotype,
    pop = pop,
    peak_activity = mean(
      sum_count_30, 
      na.rm = TRUE
    ), 
    sd_pop = sd(
      sum_count_30,
      na.rm = TRUE
    )
  ) %>%
  distinct()

```

```{r plot_peak_activity}
# per monitor
ggplot(
  data = peak_activity,  # data defined above 
  mapping = aes(
    x = pop, 
    y = peak_activity, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = peak_activity - sd_pop, 
      ymax = peak_activity + sd_pop
    ), 
    width = 0.2, 
    position = position_dodge(0.4)
  ) +
  geom_point(
    position = position_dodge(0.4)
  ) + 
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(time), 
    rows = vars(bloc)
  ) +
  TIME_theme

```

```{r dawn_dusk_sum}

# for rolling averages
peak_activity_gt_roll <- peak_activity %>%
 # filter(
#      time == '10 hours' | time == '21.5 hours'
#  ) %>%
   group_by(
    time, bloc, genotype
  ) %>%
  reframe(
    genotype = genotype,
    peak_activity_mean = mean(peak_activity, na.rm = TRUE), 
    sd_gt = sd(peak_activity, na.rm = TRUE)
  ) %>%
  distinct()


```

```{r dawndusk_plot}

# per genotype
ggplot(
  data = peak_activity_gt_roll, 
  mapping = aes(
    x = genotype, 
    y = peak_activity_mean, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = peak_activity_mean - sd_gt, 
      ymax = peak_activity_mean + sd_gt
    ), 
    width = 0.2, 
    position = position_dodge(0.4)
  ) +
  geom_point(
    position = position_dodge(0.4), 
    size = 3
  ) + 
  ylab(
    "Mean Activity"
  ) +
  scale_color_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(time), 
    rows = vars(bloc)
  ) +
  theme_light()
  
  
  

```

```{r dawn_dusk_lm}

head(diurnal_activity_roll_chamber)

#mod_data <- dawn_activity_chamber %>%
#  filter(time_interval > "10 hours" )

mod_data <- diurnal_activity_roll_chamber %>%
  filter(light == 1)

for (i in c("bloc","genotype", "pop", "chamber")) {
  mod_data[[i]] <- as.factor(mod_data[[i]])
}

mod_data[["chamber"]] <- as.factor(mod_data[["chamber"]])
str(mod_data)

activity_mod <- lme4::lmer(
  formula = mean_window ~   bloc + genotype + pop + (1|chamber),
  data = mod_data
)
mod_aov <- car::Anova(activity_mod)
mod_aov

mod_emm <- emmeans::emmeans(
  activity_mod, 
  pairwise ~ genotype, 
  adjust = "tukey"
)
mod_emm


##-----------------------------------------


```
```{r plot_model_means}

model_emm <- tibble::as_tibble(mod_emm$emmeans)
str(model_emm)

class(model_emm$genotype)

ggplot2::ggplot(
  data = model_emm, 
  mapping = aes(
    x = genotype, 
    y = emmean, 
    colour = genotype
  )
) + 
  geom_point(
    size = 3
  ) +
  geom_errorbar(
    mapping = aes(
      ymax = upper.CL, 
      ymin = lower.CL
    ), 
    width = 0.5
  ) +
  ylab(
    "Model Calibrated Estimates of \n Mean Daytime Activity (per minute)"
  ) +
  xlab(
    "Genotype"
  ) +
  scale_colour_manual(
   values = TIME_palette 
  ) +
  TIME_theme
```

## DEPRECIATED Activity plots

-   by monitor

-   by genotype

```{r monitor_plots_depricated}

# for grouped mean
ggplot2::ggplot(
  data = day2_monitor_mean, 
  mapping = aes(
    x = time_interval, 
    y = mean_window, 
    linetype = as.factor(pop)
  )
) + geom_line(
  mapping = aes(
    colour = as.factor(pop)
  )
) +
  ylab(
    "Blocked Mean of Counts (30 min windows)"
  ) #+
  scale_colour_manual(
    values = TIME_palette
  )

# for rolling mean
ggplot2::ggplot(
  data = day2_monitor_rollmean, 
  mapping = aes(
    x = time, 
    y = mean_window, 
    linetype = as.factor(pop)
  )
) + geom_line(
  mapping = aes(
    colour = genotype
  )
) +
  ylab(
    "Rolling Mean of Counts (30 min windows)"
  ) +
  scale_colour_manual(
    values = TIME_palette
  )

```

```{r plotting_gt_activity_depricated}

# for block means
ggplot2::ggplot(
  data = day2_activity_gt, 
  mapping = aes(
    x = time_interval, 
    y = mean_window, 
    colour = genotype
  )
) + 
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    mapping = aes(
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    mapping = aes(
      colour = genotype
    ), 
    #position = position_dodge2(50)
) +
  geom_point(
    mapping = aes(
      y = mean_window, 
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    )
  ) + theme_minimal()


# for rolling means
ggplot2::ggplot(
) +
  geom_rect(
    data = light_timings,
    mapping = aes(
      ymin = -Inf,
      ymax = Inf, 
      xmin = xstart,
      xmax = xend,
      alpha = as.factor(light)
    ), 
    show.legend = FALSE
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.1)
  ) +
  #geom_errorbar(
  #  mapping = aes(
  #    ymin = mean_window - sd_gt, 
  #    ymax = mean_window + sd_gt
  #  ), 
  #  width = 360, 
  #  #position = position_dodge2(50)
  #) +
  geom_ribbon(
    data = day2_activity_gt_roll,
    mapping = aes(
      x = time,
      ymin = mean_window - sd_gt, 
      ymax = mean_window + sd_gt, 
      fill = genotype
    ), 
    linetype = "dashed", 
    alpha = 0.2
  ) + 
  geom_line(
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    )
) +
  geom_point(
    data = day2_activity_gt_roll, 
    mapping = aes(
      y = mean_window, 
      x = time,
      colour = genotype
    ), 
    size = 2, 
    #position = position_dodge2(50)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_x_time(
    breaks = c(
      hours
    )
  ) 
  facet_grid(
    rows = vars(bloc)
  )

```
