---
title: "time_activity_assay"
author: "FiG-T"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: yes
    toc: yes
toc: TRUE
editor_options:
  markdown:
    wrap: 80
  chunk_output_type: inline
---

## Importing Data from Drosophila Activity Monitors

Drosophila Activity Monitors (a.k.a DAMs) (TriKinetics) are a valuable tool for
assessing the patterns of movement in fly studies. Derived metrics (peak/average
activity, number of sleeps etc. etc) are frequently used as a proxy for
"fitness". One advantage of the DAMs is that each monitor can simultaneously
measure up to 32 flies, and multiple monitors can be used at once. DAMs are thus
suitable for scaling for use in high-throughput studies.

One bottleneck in this process is the importing of data into programs capable of
performing statistical analysis. Presently this is done by copy and pasting
blocks of data into a template excel sheet. As each monitor has a separate
output file this has to be done for each monitor. This looks to be rather dull,
error prone, and time consuming.

The following script is my attempt to automate this process.

> **The initial sections of this code show the workings of how this is achieved.
> If you wish to immediately import data please run the required libraries then
> skip to [The \`import_DAMs\` function] .**

### Required Libraries

```{r libraries}

library(stringr)
library(readxl)
library(lubridate)
library(dplyr)

```

### Identifying folders and DAMs output files

This section of script sets the directory in which the folders containing the
monitor outputs are kept. This assumes that the files are organised in such a
way that the monitor output files are nested within a folder for each bloc, and
these folders are further nested in a single folder themselves.

The pattern here can be used to match the desired folders (and must be a regular
expression).

```{r list_files}

# overall directory
activity_path <- "~/OneDrive - University College London/data/time_fitness/activity"

DAMS_blocks <- list.files(
  path = activity_path, 
  pattern = "DAMS_b.*"  # the regex that matches the folder with monitor data 
)

# intiate a list
monitor_list <- list() 

# for every folder in the overall directory
for (i in DAMS_blocks){
  
  monitor_list[[i]] <- list.files(
    path = paste(
    activity_path, 
    i, 
    sep = "/"
    )
  )
  
  message(
    paste(
      "Monitors from", 
      i,
      "added to list."
    )
  )
}

```

#### The Metafile

The approach used here identified all relevant folders and output files (above)
and combines them (below) with associated data from a provided metafile. The
following chunk is used to import this metafile into R.

The metadata file must have information on the date of the experiment (in day -
month - year format) and a "bloc" column - the information here should match (or
partially match if numeric) the directory names.

```{r monitor_metafile}

# read in metadata file 
monitor_info <- read_excel(
  path = paste0(
    activity_path, 
    "/", 
    list.files(
      path = activity_path, 
      pattern = ".*meta.*"
    )
  ), 
  col_names = TRUE
)

# convert to date type
monitor_info$date <- dmy(monitor_info$date)

# determine the number of blocs 
blocs <- unique(monitor_info$bloc)

# initiate list
monitor_included <- list()

for (j in blocs) {
  
  filter_table <- monitor_info %>% 
    filter(
      bloc == j
    ) 
  
  # separate each monitor in each bloc into a separate variable in a list
  monitor_included[[j]] <- unique(filter_table$monitor)
  
  monitor_included[[j]] <- str_c(
    "Monitor", monitor_included[[j]], ".txt"
  )
  
}

# name the components in this list
names(monitor_included) <- stringr::str_c(
  "DAMS_b", blocs, sep = ""
)


```

Depending on how the data was copied into the directory, some redundant (not
used) monitor output files may be present. These are removed by filtering to
only include paths to monitors that are present in both the directory and the
metadata file.

```{r filter_included}

for( i in seq_along(monitor_list)) {
  
  # only include monitors in list that are also included on the metadata file
  monitor_list[[i]] <- monitor_list[[i]][
    monitor_list[[i]] %in% monitor_included[[i]]
  ]
  
}

# initiate list 
monitor_path_list <- list()

for (i in seq_along(monitor_list)) {
  
  bloc <- names(monitor_list)[i]
  
  bloc_path <- paste(
    activity_path, 
    bloc, 
    sep = "/"
  )
  
  #print(bloc_path)
  
  # for each monitor in a bloc
  for(k in seq_along(monitor_list[[bloc]])) {
    
    monitor <- monitor_list[[bloc]][k]
    
    monitor_path <- paste(
      bloc_path, 
      monitor, 
      sep = "/"
    )
    
    monitor_path_list[[bloc]][k] <- monitor_path
    
    #print(monitor_path)
  
  }
}

```

Once the paths to all of the relevant monitors have been constructed, the data
from each of them can be imported and merged with the associated data in the
provided metafile.

Note: this method will only keep data from equal or after the date provided in
the metadata.

```{r test_build}

# to test: 
#monitor_path_list$DAMS_b2 <- monitor_path_list$DAMS_b2[1] # only take the first file
#monitor_list_test <- list(monitor_list$DAMS_b2[1])
#names(monitor_list_test) <- "DAMS_b2"

monitor_list <- monitor_list[
  names(monitor_list) %in% names(monitor_path_list) == TRUE
]
 

for (i in seq_along(monitor_path_list)) {
  
  path_lists <- monitor_path_list[[i]]
  
  monitor <- monitor_list[[i]]
  
  bloc <- str_replace_all(
      string = names(monitor_path_list[i]), 
      pattern = "[:alpha:]|[:punct:]", 
      replacement = ""
  )
  
  for (k in seq_along(path_lists)) {
    
    tmp_table <- read.delim(
      file = path_lists[k], 
      header = FALSE
    )
    
    tmp_table <- tmp_table[,c(1:4,10:42)]
    
    names(tmp_table) <- c(
      "index", "date", "time", "status", "light", 
      str_c(
        "chamber_", 
        c(1:32), 
        sep = ""
      )
    )
    print(monitor[k])
    
    tmp_table$monitor <- str_replace_all(
      string = monitor[k] , 
      pattern = "[:alpha:]|[:punct:]", 
      replacement = ""
    )
    
    tmp_table$monitor <- as.numeric(tmp_table$monitor)
    
    tmp_table$bloc <- as.numeric(bloc)
    
    tmp_table$date <- dmy(tmp_table$date)
    
    #print(tail(tmp_table))
    
    
    tmp_table <- left_join(
      x = tmp_table, 
      y = monitor_info, 
      by = join_by(
        monitor == "monitor", 
        bloc == "bloc"
      )
    )
    
    start_date <- unique(tmp_table$date.y)
    
    tmp_table <- tmp_table %>%
      filter(
        date.x >= unique(tmp_table$date.y)
      )
    
    #print(tmp_table)
    
    if (i == 1 & k == 1) {
      dams_table <- tmp_table
    } else {
      dams_table <- rbind(
        dams_table,
        tmp_table
      )
    }
    
  }
  
}

dams_table <- dams_table %>%
  select(
    index:light, monitor, bloc, genotype, pop, 
    chamber_1:chamber_32
  )

dams_table <- dams_table %>%
  select(
    index:light, names(monitor_info), 
    chamber_1:chamber_32
  )



names(dams_table)[2] <- "date"
```

## The \`import_DAMs\` function

To make this script applicable for regular use the following function has been
complied. Please note that this was written based on the personal file system
structure that I use and may therefore be temperamental is your system does not
match this. For this reason I have also include a breakdown of the code above.

### File system structure 

Two types of files are required for this function to work.

1.  **A folder (or folders) containing the raw activity files.**

    These folder names should all have the same stem + bloc/timepoint number (if
    using). All the raw data files should be in the standard DAMS output format
    (e.g. Monitor1.txt)

2.  **A metadata file with the biological information for each monitor in each
    folder.**

#### Metdata file format

The metadata file should be an excel file with information on:

-   the monitor number

-   the bloc

-   the date o*f the start* of the experiment (i.e: when they were added to the
    chambers) - this is used to automatically calculate the burnin time.

-   the genotype

-   the population (or other unique identifier)

E.g.:

| monitor | bloc | date       | genotype | pop |
|---------|------|------------|----------|-----|
| 1       | 1    | 01.01.2024 | A        | 1   |
| 1       | 2    | 10.10.2024 | B        | 3   |

: Example metadata table

In this instance the folder names containing the raw data must end with 1 and 2
(for bloc 1 and 2 respectively).

E.g:

Within the directory: *\~/files/data/project/activity* You should have...

*DAMS_Bloc1* ( a folder containing *Monitor1.txt*)

*DAMS_Bloc2* (a folder containing *Monitor1.txt)*

activity_metadata.xlsx (the example table written out above)

```{r import_DAMs}

import_DAMs <- function(
        path,  # the path to the general directory
        directory_pattern = "DAMS_b.*", # the regex that matches the folder with monitor data 
        meta_pattern = ".*meta.*",  # the regex that uniquely matches the file with metadata
        meta_col_1 = "monitor", # the column names in the metadata file
        meta_col_2 = "bloc",
        meta_date_col = 3,  # numeric indicator of where the date values are in the metafile
        bloc_name = "DAMS_b", # the link between the metadata bloc value and folder name.
        numeric_bloc = TRUE
){
  
  DAMS_blocks <- list.files(
    path = path, 
    pattern = directory_pattern  
  )
  
  # intiate a list
  monitor_list <- list() 
  
  for (i in DAMS_blocks){
    monitor_list[[i]] <- list.files(
      path = paste(
        path, 
        i, 
        sep = "/"
      )
    )
    
    # report back on progress
    message(
      paste(
        "Monitors from", 
        i,
        "added to list."
      )
    )
  }
  
  # read in metadata file 
  monitor_info <- read_excel(
    path = paste0(
      path, 
      "/", 
      list.files(
        path = path, 
        pattern = meta_pattern
      )
    ), 
    col_names = TRUE
  )
  
  message(
    paste(
        "Metadata from ", 
        list.files(
          path = path, 
          pattern = meta_pattern
        ),
        " successfully imported."
      )
  )
  
  # convert to date type
  monitor_info$date <- dmy(monitor_info$date)

  # determine the number of blocs 
  blocs <- unique(monitor_info$bloc)
  
  # initiate list
  monitor_included <- list()

  for (j in blocs) {
  
    filter_table <- monitor_info %>% 
      filter(
        bloc == j
      ) 
  
    # separate each monitor in each bloc into a separate variable in a list
    monitor_included[[j]] <- unique(filter_table$monitor)
  
    monitor_included[[j]] <- str_c(
      "Monitor", monitor_included[[j]], ".txt"
    )
  }
  
  # name the components in this list
  names(monitor_included) <- stringr::str_c(
    bloc_name, blocs, sep = ""
  )
  
  for( i in seq_along(monitor_list)) {
  
    # only include monitors in list that are also included on the metadata file
    monitor_list[[i]] <- monitor_list[[i]][
      monitor_list[[i]] %in% monitor_included[[i]]
    ]
  }
  
  # initiate list 
  monitor_path_list <- list()
  
  # import and combine activity and metadata
  for (i in seq_along(monitor_list)) {
  
    bloc <- names(monitor_list)[i]
  
    bloc_path <- paste(
      path, 
      bloc, 
      sep = "/"
    )
  
    # for each monitor in a bloc
    for(k in seq_along(monitor_list[[bloc]])) {
    
      monitor <- monitor_list[[bloc]][k]
    
      monitor_path <- paste(
        bloc_path, 
        monitor, 
        sep = "/"
      )
    
      monitor_path_list[[bloc]][k] <- monitor_path
    }
  }
  
  # remove monitors that are not included in the list of paths
  monitor_list <- monitor_list[
    names(monitor_list) %in% names(monitor_path_list) == TRUE
  ]
  
  # initiate loop through directories containing monitor files
  for (i in seq_along(monitor_path_list)) {
  
    path_lists <- monitor_path_list[[i]]
  
    monitor <- monitor_list[[i]]
    
    if (numeric_bloc == TRUE){
      # remove letters and punction from folder name
      bloc <- str_replace_all(
        string = names(monitor_path_list[i]), 
        pattern = "[:alpha:]|[:punct:]", 
        replacement = ""
      )
    } else {
      bloc <- names(monitor_path_list[i])
    }
    
    # loop through paths within a folder
    for (k in seq_along(path_lists)) {
    
      tmp_table <- read.delim(
        file = path_lists[k], 
        header = FALSE
      )
    
      tmp_table <- tmp_table[,c(1:4,10:42)]
    
      names(tmp_table) <- c(
        "index", "date", "time", "status", "light", 
        str_c(
          "chamber_", 
          c(1:32), 
          sep = ""
        )
      )
    
      message(
        paste(
          "Activity data imported from ", 
          path_lists[k]
        )
      )
      
      tmp_table$monitor <- str_replace_all(
        string = monitor[k] , 
        pattern = "[:alpha:]|[:punct:]", 
        replacement = ""
      )
        
      tmp_table$monitor <- as.numeric(tmp_table$monitor)
       
      if (numeric_bloc == TRUE){
        tmp_table$bloc <- as.numeric(bloc)
      }
    
      tmp_table$date <- dmy(tmp_table$date)
      
      # combine with metadata

      tmp_table <- left_join(
        x = tmp_table, 
        y = monitor_info, 
        by = join_by(
          monitor == "monitor", 
          bloc == "bloc"
        )
      )
      
      # start date taken from metafile 
      start_date <- unique(tmp_table$date.y)
    
      tmp_table <- tmp_table %>%
        filter(
          date.x >= unique(tmp_table$date.y)
        )
      
      message(
        paste(
          "Metadata successfully combined with activity data"
        )
      )
      
      if (i == 1 & k == 1) {
        dams_table <- tmp_table
        
        message(
          "DAMs table created"
        )
      } else {
        dams_table <- rbind(
          dams_table,
          tmp_table
        )
        
        message(
          "Information added to DAMs table"
        )
      }
      
    } # close loop through file
  }   # close loop through folders
  
  dams_table <- dams_table %>%
    select(
      index:light, names(monitor_info)[-meta_date_col], 
      chamber_1:chamber_32
    )
  
  names(dams_table)[2] <- "date"
   
  message(
    "--------------------------------------------------------------------------------
    --------------------------------------------------------------------------------
    Operation complete - all files imported successfully
    --------------------------------------------------------------------------------
    --------------------------------------------------------------------------------"
  )
  return(dams_table)   
  
} # close function


```

> > ------ Make sure that the metadata file has been updated ------

```{r running_import_DAMs}

dams_table <- ""

dams_table <- import_DAMs(
  path = "~/OneDrive - University College London/data/time_fitness/activity", # the path to the general directory
  directory_pattern = "DAMS_b.*", # the pattern that matches the folders with raw data 
  meta_pattern = ".*meta.*", # the regex to find your metadata file
  meta_date_col = 3,  # the column containing the date information 
  bloc_name = "DAMS_b",  # the link between your folder names and bloc timepoints... e.g. In the above example the blocs are 1 and 2 but the folders are DAMS_Bloc1 and DAMS_Bloc2. The connector is therefore "DAMS_Bloc"
  numeric_bloc = TRUE
)

openxlsx::write.xlsx(
  x = dams_table, 
  file = "~/OneDrive - University College London/data/time_fitness/activity/time_activity_data.xlsx"
)
```
