---
title: "T.I.M.E Productivity Assay"
author: "FiG-T"
date: "`r Sys.Date()`"
output: 
html_document: 
  keep_md: yes
    toc: yes
  toc: TRUE
  editor_options:
  markdown:
    wrap: 80
  chunk_output_type: inline
---

# Introduction

This markdown is for the "productivity" assay, included as part of the T.I.M.E fitness assays on experimentally evolving *Drosophila melanogaster* populations.

> Productivity is considered here to be:
>
> *The number of adult offspring arising from a single female.*

# Libraries 

```{r libraries}
library(dplyr)
library(readxl)
library(tidyr)
```

# Palettes

```{r theme_setup}
# set the palette
TIME_palette <- c(
  mX = "orchid1", 
  tX = "darkorchid", 
  mM = "deepskyblue2", 
  tT = "firebrick3"
)

col2rgb(TIME_palette)

TIME_theme <- theme_light(
) +
  theme(
    strip.background = element_rect(
      fill = "snow4"
    )
  )
```

# Importing data 

Raw data can be imported directly from the excel file. This has information on:

-   the genotype (mX, tX, mM, tT)

-   the population (1-14)

-   the subsample (1_30)

-   the timepoint

-   the number of adults recorded

```{r import_data}
# import data 
prod_data <- readxl::read_excel(
  path = "~/OneDrive - University College London/data/time_fitness/production/time_production_raw_data.xlsx", 
  sheet = "productivity", 
  col_names = TRUE
)

# remove notes column
prod_data <- prod_data %>%
  select(
    genotype:production
  )

# rename production column 
names(prod_data)[5] <- "n_adults"
```

# Data processing 

```{r format_data}
# set levels for genotype 
prod_data$genotype <- factor(
  prod_data$genotype, 
  levels = c("mM", "mX", "tX", "tT")
  )
# set levels for pop
prod_data$pop <- factor(
  prod_data$pop, 
  levels = c("11",
             "12","13",
             "1", "2", "3", "4", "5" , "6", "7", "8", "9", "10",
             "14", "15", "16")
  )


```

The data can be grouped by population and/or genotype, along with timepoint.

To get an overview of the data:

```{r overview_plot}
# to create the boxplot - used to spot any obvious errors in the data 
ggplot2::ggplot(
  data = prod_data, 
  mapping = aes(
    x = pop, 
    y = n_adults,
    fill = genotype
  )
) +
  geom_boxplot(
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  facet_grid(
    cols = vars(timepoint)
  ) +
  TIME_theme

```

## Population averages

```{r population_averages}

# summarise per population per genotype
prod_mean_pop <- prod_data %>% 
  group_by(
    pop, timepoint
  ) %>%
  reframe(
    genotype = genotype, 
    pop = pop, 
    timepoint = timepoint, 
    mean_prod = mean(n_adults, na.rm = TRUE), 
    sd_prod = sd(n_adults, na.rm = TRUE), 
    N = n(), 
    se_prod = sd_prod/sqrt(N)
  ) %>%
  distinct()
```

```{r plot_pop_averages, warning=FALSE, message=FALSE}

# group by timepoint, plot pop summaries 
ggplot2::ggplot(
  data = prod_mean_pop, 
  mapping = aes(
    x = pop, 
    y = mean_prod, 
    colour = genotype
  )
) + 
  geom_errorbar(
    mapping = aes(
      ymin = mean_prod - sd_prod, 
      ymax = mean_prod + sd_prod
    ), 
    alpha = 0.5, 
    width = 0.3
  ) +
  geom_point(
    size = 5
  ) +
  ylab(
    "Number of Adult Offspring"
  ) +
  facet_grid(
    cols = vars(timepoint)
  ) +
  scale_colour_manual(
    values = TIME_palette
  ) +
  TIME_theme
```

```{r save_pop_plots}
# to save the above plot
ggsave(
  filename = "/Users/finleythomas/Library/CloudStorage/OneDrive-UniversityCollegeLondon/Results/time_fitness/time_productivity_pop_means_06_2024.png", 
  plot = last_plot(),
  width = 9, 
  height = 6, 
  bg = "transparent"
)
```

```{r line_plots}

# to create the line plot:
ggplot2::ggplot(
  data = prod_mean_pop, 
  mapping = aes(
    x = timepoint, 
    y = mean_prod, 
    colour = genotype, 
    fill = genotype, 
    linetype = pop
  )
) + 
  geom_errorbar(
    mapping = aes(
      ymin = mean_prod - sd_prod, 
      ymax = mean_prod + sd_prod
    ), 
    alpha = 0.5, 
    width = 0.3, 
    position = position_dodge(0.4), 
    show.legend = FALSE
  ) +
  geom_point(
    size = 3, 
    position = position_dodge(0.4)
  ) +
  geom_line(
    position = position_dodge(0.4), 
    show.legend = FALSE
  ) + 
  scale_colour_manual(
    values = TIME_palette
  ) +
  scale_fill_manual(
    values = TIME_palette
  ) +
  scale_linetype_manual(
    values = c(
      "solid", "solid", "solid",
      "dashed", "dashed", "dashed","dashed", "dashed",  
      "dashed", "dashed", "dashed","dashed", "dashed", 
      "solid", "solid", "solid"
    )
  ) +
  TIME_theme
```

```{r save_line_plots}
ggsave(
  filename = "/Users/finleythomas/Library/CloudStorage/OneDrive-UniversityCollegeLondon/Results/time_fitness/time_productivity_pop_lines_06_2024.png", 
  plot = last_plot(),
  width = 9, 
  height = 6, 
  bg = "transparent"
)
```

## Genotype averages

```{r genotype_averages}

prod_mean_gt <- prod_mean_pop %>% 
  group_by(
    genotype, timepoint
  ) %>%
  reframe(
    genotype = genotype, 
    timepoint = timepoint, 
    mean_prod_gt = mean(mean_prod, na.rm = TRUE), 
    sd_prod = sd(mean_prod, na.rm = TRUE), 
    N = n(), 
    se_prod = sd_prod/sqrt(N)
  ) %>%
  distinct()

```

```{r plot_genotype_averages}

ggplot2::ggplot(
  data = prod_mean_gt, 
  mapping = aes(
    x = genotype, 
    y = mean_prod_gt, 
    colour = genotype
  )
) +
  geom_errorbar(
    mapping = aes(
      ymin = mean_prod_gt - sd_prod, 
      ymax = mean_prod_gt + sd_prod
    ), 
    width = 0.3
  ) +
  geom_point(
    size = 3
  ) +
  ylab(
    "Number of Adult Offspring (per female)"
  ) +
  scale_alpha_manual(
    values = c(1, 0.5)
  ) + 
  scale_colour_manual(
    values = TIME_palette
  ) + 
  facet_grid(
    cols = vars(timepoint)
  ) +
  TIME_theme

```

```{r save_gt_summary_plot}

ggsave(
  filename = "/Users/finleythomas/Library/CloudStorage/OneDrive-UniversityCollegeLondon/Results/time_fitness/time_productivity_gt_means_06_2024.png", 
  plot = last_plot(),
  width = 9, 
  height = 8, 
  bg = "transparent"
)

```

## Summary Stats

```{r prod_mod}

prod_mod <- lme4::lmer(
  data = prod_data, 
  formula = n_adults ~ genotype + pop + timepoint + (1|replicate)
)


prod_aov <- car::Anova(prod_mod)
prod_aov

anova(prod_mod)

prod_emm <- emmeans::emmeans(
  prod_mod, 
  pairwise ~ genotype + timepoint , 
  adjust = "tukey"
)
prod_emm

prod_contrasts <- tibble::as_tibble(prod_emm$contrasts)
names(prod_contrasts)

prod_contrasts %>%
  filter(
    p.value <= 0.05
  ) %>%
  select(
    contrast, p.value
  )
```
